SAT Solving for Vesicle Traffic Systems in Cells

### 2017-07-20 paper/encoding.tex
add

\begin{df}
  A VTS $G$ is tuple $(\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$, where
  \begin{itemize}
  \item $\nodes$ is a set of nodes representing compartments in the VTS,
  \item $\mols$ is the set of molecules flowing in the system, 
  \item $\nlabel : \nodes \maps \powerset{\mols}$ defines the molecules present in the nodes  
  \item $ \nodes \times (\powerset{\mols}-\emptyset) \times \nodes \subseteq \edges$ is the
    set of edges with molecule sets as labels
  \item $\pairs \subseteq \mols \times \mols$ is pairing relation
  \item $\nodef : \mols \maps \powerset{\mols} \maps \booleans$ is activity maps for nodes
  \item $\edgef : \mols \maps \powerset{\mols} \maps \booleans $ is activity maps for edges
  \end{itemize}
\end{df}
$\pairs$ defines which molecules can fuse with which molecules.
%
Let $\pairs(M')$ denote $\{m|(m,m') \in P \text{ and } m' \in M'\}$.
%
$\nodef$ and $\edgef$ are used to define activity of molecules on
nodes and edges respectively.
%
A molecule $k$ is {\em active} at node $n$ if $k \in \nlabel(n)$ and
$\nodef(k,\nlabel(n))$ is true.
%
A molecule $k$ is {\em active} at edges $(n,M',n')$ if $k \in M'$ and
$\edgef(k,M')$ is true.
%
We call $G$ {\em well-structured} if molecules $M$ is divided into
two equal partitions $M_1$ and $M_2$ such that $P \subseteq M_1 \times M_2$, and
for each $(n,M',n') \in \edges$, $n \neq n'$, 
$M' \subseteq \nlabel(n)$, and $M' \subseteq \nlabel(n')$.
%

@@

A {\em path} in $G$ is a sequence $n_1,...,n_\ell$ of nodes 
such that $(n_i,\_,n_{i+1}) \in \edges$ for each $ 0 < i < \ell$.
%
For a molecule $m \in M$,
an {\em $m$-path} in $G$ is a sequence $n_1,...,n_\ell$ of nodes 
such that $(n_i,M',n_{i+1}) \in \edges$ and $m \in M'$ for
each $ 0 < i < \ell$.
%
A node $n'$ is {\em ($m$-)reachable} from node $n$ in $G$ if there is a ($m$-)path
$n,...,n'$ in $G$.
%
% A node $n'$ is {\em $m$-reachable} from node $n$ in $G$ if there is a
% $m$-path $n,...,n'$ in $G$.
%
We call $G$ {\em stable} if for each $(n,M',n') \in \edges$ and $m \in M'$,
$n$ is $m$-reachable from $n'$.
%
An edge $(n,M',n') \in \edges$ {\em fuses} with its destination node $n'$
if there are molecules $m,m' \in \mols$ such that $m$ is active in
$(n,M',n')$, $m'$ is active in $n'$, and $(m,m') \in \pairs$.
%
We call $G$ {\em well-fused} if each edge $(n,M',n') \in \edges$ fuses
with non-empty fusing molecules $M'' \subseteq M'$
and $\pairs(M'')$ are not active in any other node.
%
We call $G$ {\em connected} if for each $n,n' \in \nodes$,
$n'$ is reachable from $n$ in $G$.
%
We call $G$ $k$-connected if for each $\edges' \subseteq \edges$ and $|\edges'| < k$,
VTS $(\nodes,\mols,\edges-E',\nlabel,\pairs,\edgef,\nodef)$ is connected.

@@

\item Boolean variable $p_{k,k'}$ indicates if $(m,m') \in \pairs$
\item uninterpreted Boolean functions $\nodef_m : \booleans^\mu \maps \booleans$
encoding $\nodef(m)$ map
\item uninterpreted Boolean functions $\edgef_m : \booleans^\mu \maps \booleans$
encoding $\edgef(m)$ map
\end{enumerate}
We also have auxiliary Boolean variables that will help us encode the well-fused and stability properties 
\begin{enumerate}
\item $a_{i,m}$ indicates that molecule $m$ is active at node $i$, i.e., $\nodef(m,L(i))$
  holds
\item $b_{i,j,q,m}$ indicates that molecule $m$ is active at $q$th edge $(i,M',j)$ between $i$ and $j$, i.e., $\edgef(m,M')$ holds
\item $r_{i,j,m,p}$ indicates that there is a $m$-path from
  $i$ to $j$ of length less than or equal to $p$
\end{enumerate}
For $k$-connected property, we also use the following auxiliary Boolean variables
\begin{enumerate}
\item $d_{i,j,q}$ indicates $q$th edge between $i$ and $j$ is dropped
\item $r'_{i,j}$ indicates that there is a path from $i$ to $j$ in the modified VTS
\end{enumerate}

@@

% \subsubsection{Constraints on the structure of the edges, nodes and }
\subsubsection{Constraints on presence, activity of the molecule, and pairing matrix}
%
We need the following constraints~\eqref{eq:f0} and~\eqref{eq:c3}
to encode the basic structure of VTSs.
%
For an edge to exist it should have one molecule present. 
%
\begin{align}
  \bigwedge\limits_{i,j,q} (\bigvee_m e_{i,j,q,m} )\limplies e_{i,j,q}\tag{V1}\label{eq:f0}
\end{align}
If a molecule is active on an edge, it should be present on the edge.
%
\begin{align}
  \bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies e_{i,j,q,m}\tag{V2}\label{eq:f1}
\end{align}
A molecule should be present to be active on a node.  
\begin{align}
  \bigwedge\limits_{i,m} a_{i,m} \limplies n_{i,m}
  \tag{V3}\label{eq:c4}
\end{align}
The edge labels are subset of the node labels of the source and target nodes.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies (n_{i,m} \land n_{j,m} )\tag{V4}\label{eq:c0}
\end{align}
Self edges are not allowed. 
\begin{align}
   \bigwedge\limits_{i,q} \neg e_{i,i,q}\tag{V5}\label{eq:c2}
\end{align}
We fix first half as Q-SNAREs and rest as R-SNAREs and set diagonal blocks in pairing matrix to be 0's.
%It reduces the search space.
% Condition on $p_{kk'}$. \ashu{why???} .
\begin{align}
  \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y)
  \tag{V6}\label{eq:c3}
\end{align}

@@

We have encoded the following variants of VTSs.
%
The variants are due to different combinations of constraints on the
activity on nodes and edges.
%
\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\end{enumerate}
%
% For the version A, every present molecule is in active state.
%
In the table~\ref{tab:grph}, we present the constraints involved in each version.
%
The column two of the table shows that the constraints V1-V8, R1-R2, and D1-D4 are present
for every variant.
%
The column three of the table lists the constraints that are different among the variants.
%
%
One of the restriction is where the activity of the present molecule is dependent on the presence of the other molecules.
%
For example in version B, D, F activity of the molecules on the node is a boolean function of the presence of other molecules on that node; Anb.
%
Similarly for the case of the edge in version C and D; Aeb.
%
In the case of versions F the activity of the molecules on the edges
is described by pairing matrix; Aep.

>>>>

\begin{df}
  A VTS $G$ is tuple $(\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$, where
  \begin{itemize}
  \item $\nodes$ is a set of nodes representing compartments in the VTS,
  \item $\mols$ is the set of molecules flowing in the system, 
  \item $\nlabel : \nodes \maps \powerset{\mols}$ defines the molecules present in the nodes  
  \item $ \nodes \times (\powerset{\mols}-\emptyset) \times \nodes \subseteq \edges$ is the
    set of edges with molecule sets as labels
  \item $\pairs : \mols \times \mols$ is pairing relation
  \item $\nodef : \mols \maps \powerset{\mols} \maps \booleans$ is activity maps for nodes
  \item $\edgef : \mols \maps \powerset{\mols} \maps \booleans $ is activity maps for edges
  \end{itemize}
\end{df}
$\pairs$ defines which molecules can fuse with which molecules.
%
Let $\pairs(M')$ denote $\{m|(m,m') \in P \text{ and } m' \in M'\}$.
%
$\nodef$ and $\edgef$ are used to define activity of molecules on
nodes and edges respectively.
%
A molecule $k$ is {\em active} at node $n$ if $k \in \nlabel(n)$ and
$\nodef(k,\nlabel(n))$ is true.
%
A molecule $k$ is {\em active} at edges $(n,M',n')$ if $k \in M'$ and
$\edgef(k,M')$ is true.
%
We call $G$ {\em well-structured} if molecules $M$ is divided into
two equal partitions $M_1$ and $M_2$ such that $P \subseteq M_1 \times M_2$, and
for each $(n,M',n') \in \edges$, $n \neq n'$, 
$M' \subseteq \nlabel(n)$, and $M' \subseteq \nlabel(n')$.
%

@@

A {\em path} in $G$ is a sequence $n_1,...,n_\ell$ of nodes 
such that $(n_i,\_,n_{i+1}) \in \edges$ for each $ 0 < i < \ell$.
%
For a molecule $m \in M$,
an {\em $m$-path} in $G$ is a sequence $n_1,...,n_\ell$ of nodes 
such that $(n_i,M',n_{i+1}) \in \edges$ and $m \in M'$ for
each $ 0 < i < \ell$.
%
A node $n'$ is {\em reachable} from node $n$ in $G$ if there is a path
$n,...,n'$ in $G$.
%
A node $n'$ is {\em $m$-reachable} from node $n$ in $G$ if there is a
$m$-path $n,...,n'$ in $G$.
%
We call $G$ {\em stable} if for each $(n,M',n') \in \edges$ and $m \in M'$,
$n$ is $m$-reachable from $n'$.
%
An edge $(n,M',n') \in \edges$ {\em fuses} with its destination node $n'$
if there are molecules $m,m' \in \mols$ such that $m$ is active in
$(n,M',n')$, $m'$ is active in $n'$, and $(m,m') \in \pairs$.
%
We call $G$ {\em well-fused} if each edge $(n,M',n') \in \edges$ fuses
with non-empty fusing molecules $M'' \subseteq M'$
and $\pairs(M'')$ are not active in any other node.

%
We call $G$ {\em connected} if for each $n,n' \in \nodes$,
$n'$ is reachable from $n$ in $G$.
%
We call $G$ $k$-connected if for each $\edges' \subseteq \edges$ and $|\edges'| < k$,
VTS $(\nodes,\mols,\edges-E',\nlabel,\pairs,\edgef,\nodef)$ is connected.

@@

\item Boolean variable $p_{k,k'}$ indicates if $(m,m') \in \pairs$
\item uninterpreted Boolean functions $\nodef_m : \booleans^\mu \maps \booleans$
encoding $\nodef(m)$ map
\item uninterpreted Boolean functions $\edgef_m : \booleans^\mu \maps \booleans$
encoding $\edgef(m)$ map
\end{enumerate}
We also have auxiliary Boolean variables that will help us encode the well-fused and stability properties 
\begin{enumerate}
\item $a_{i,m}$ indicates that molecule $m$ is active at node $i$, i.e., $\nodef(m,L(i))$
  holds
\item $b_{i,j,q,m}$ indicates that molecule $m$ is active at $q$th edge $(i,M',j)$ between $i$ and $j$, i.e., $\edgef(m,M')$ holds
\item $r_{i,j,m,p}$ indicates that there is a path from $i$ to $j$ of length less than or equal to $p$
\end{enumerate}
For $k$-connected property, we also use the following auxiliary Boolean variables
\begin{enumerate}
\item $d_{i,j,q}$ indicates $q$th edge between $i$ and $j$ is dropped
\item $r'_{i,j}$ indicates that there is a path from $i$ to $j$ is a modified VTS
\end{enumerate}

@@

% \subsubsection{Constraints on the structure of the edges, nodes and }
\subsubsection{Constraints on presence, activity of the molecule, and pairing matrix}
%
We need the following constraints~\eqref{eq:f0} and~\eqref{eq:c3}
to encode the basic structure of VTSs.
%
For an edge to exist it should have one molecule present. 
%
\begin{align}
  \bigwedge\limits_{i,j,q} (\bigvee_m e_{i,j,q,m} )\limplies e_{i,j,q}\tag{V1}\label{eq:f0}
\end{align}
If a molecule is active on an edge, it should be present on the edge.
%
\begin{align}
  \bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies e_{i,j,q,m}\tag{V2}\label{eq:f1}
\end{align}
A molecule should be present to be active on a node.  
\begin{align}
  \bigwedge\limits_{i,m} a_{i,m} \limplies n_{i,m}
  \tag{V3}\label{eq:c4}
\end{align}

The edge labels are subset of the node label of source and target compartments.

\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies (n_{i,m} \land n_{j,m} )\tag{V4}\label{eq:c0}
\end{align}
Self edges are not allowed. 
\begin{align}
   \bigwedge\limits_{i,q} \neg e_{i,i,q}\tag{V5}\label{eq:c2}
\end{align}
We fix first half as Q-SNAREs and rest as R-SNAREs and set diagonal blocks in pairing matrix to be 0's.
%It reduces the search space.
% Condition on $p_{kk'}$. \ashu{why???} .
\begin{align}
  \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y)
  \tag{V6}\label{eq:c3}
\end{align}

@@

We have encoded the following variants of VTSs.
%
The variants are due to different combinations of constraints on the
activity of molecules on the nodes and edges.
%
\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\end{enumerate}
%
% For the version A, every present molecule is in active state.
%
In the table~\ref{tab:grph}, we present the constraints involved in each version.
%
The column two of the table shows that the constraints V1-V8, R1-R2, and D1-D4 are present
for every variant.
%
The column three of the table lists the constraints that are different among the variants.
%
%
One of the restriction is where the activity of the present molecule is dependent on the presence of the other molecules.
%
For example in version B, D, F activity of the molecules on the node is a boolean function of the presence of other molecules on that node; Anb.
%
Similarly for the case of the edge in version C and D; Aeb.
%
In the case of versions F the activity of the molecules on the edges
is described by pairing matrix; Aep.

### 2017-07-20 paper/experiments.tex
fixed-table0-intro-bio-exper
in 12 page
add

\begin{table}[t]
\centering
% \def\arraystretch{1.6}
  \begin{tabular}{|c|c|c|c|}
    \hline
    {\multirow{2}{*}{\textbf{Variant}}}  & 
    \multicolumn{2}{c|}{\textbf{Constraints}} & 
    \textbf{Graph connectivity}
    % \multicolumn{2}{c|}{\textbf{Graph connectivity}}
    \\
   \cline{2-4}
   &  \textbf{Rest} & \textbf{Activity} & \textbf{\textbf{Least connected}} % & \textbf{Sampling: Guarantee}
\\ \hline

@@

We have implemented the encodings for each variants
using the python interface of $\zthree$ in a tool.
%
Our tool allows the user to choose a model and the size of the network
besides other parameters like connectivity and number of parallel
edges.
%
It uses Z3 Python interface to build the needed constraints and
applies $\zthree$ solver on the constraints to find a model (
a satisfying assignment that respects the constraints).
%
This tool also translates the satisfying model found by $\zthree$ into
a VTS and presents a visual output to the user in form of annotated graph.
%
% The output graph will satisfy the underlying rules of the system, so
% it always is a valid vesicle transport network.
%
% The resultant graph provides an overview of the underlying VTS
% where directed labeled edges and nodes provide the
% complete information about the system.
%
We also visually report the dropped edges to make the graph
disconnected, which gives information about the connectivity of the
graph.

@@

In Table~\ref{tab:qf-grabh}, we present the running times for the search of
VTSs of sizes 2 to 10 that satisfy the variants.
%
The experiments were done on a machine with Intel(R) Core(TM) i3-4030U
CPU @ 1.90GHz processor and 4GB RAM.
%
We have compared our performance with the performance of our earlier 
CBMC based implementation.
%
For example, the formula for variation F, the total number of
compartments ($|N|$) equals to 10, returns in 129.78 minutes (7786.8 secs)
with a SAT result.
%
In comparison, CBMC results in OUT OF MEMORY for $|N|$ greater than 5.
%
``!'' indicate that the constraints were unsatisfiable.
%
Hence with the use of this novel encoding, we are able to scale the
system to a much larger compartmentalized system, especially to
eukaryotic cells with a total number of ten compartments.
%
% \ankit{
% This additional scalability can provide us leverage to ask for related
% open questions that were not previously possible, for example, what is
% the minimum number of molecules required to satisfy the vesicular
% traffic system in eukaryotes.}
%
Furthermore, we experimented with limits of our tools and found
that $\zthree$ was able to solve the constraints up to $\sim{14-18}$ nodes.

>>>>

\begin{table}[t]
\centering
% \def\arraystretch{1.6}
  \begin{tabular}{|c|c|c|c|}
    \hline
    {\multirow{2}{*}{\textbf{Variant}}}  & 
    \multicolumn{2}{c|}{\textbf{Constraints}} & 
    {\multirow{2}{*}{\textbf{Graph connectivity}}}
    % \multicolumn{2}{c|}{\textbf{Graph connectivity}}
    \\
   \cline{2-3}
   &  \textbf{Rest} & \textbf{Activity} &  % & \textbf{Sampling: Guarantee}
\\ \hline

@@

We have implemented the encodings for each variants
using the python interface of $\zthree$ in a tool.
%
Our tool allows the user to choose a model and the size of the network
besides other parameters like connectivity and number of parallel
edges.
%
It uses Z3 Python interface to build the needed constraints and
applies $\zthree$ solver on the constraints to find a model (
a satisfying assignment that respects the constraints).
%
This tool also translates the satisfying model found by $\zthree$ into
a VTS and presents a visual output to the user in form of annotated graph.
%
% The output graph will satisfy the underlying rules of the system, so
% it always is a valid vesicle transport network.
%
% The resultant graph provides an overview of the underlying VTS
% where directed labeled edges and nodes provide the
% complete information about the system.
%
We also visually report the dropped edges required to disconnect the graph, it gives information about the connectivity of the graph.

@@

In Table~\ref{tab:qf-grabh}, we present the running times for the search of
VTSs of sizes 2 to 10 that satisfy the variants.
%
The experiments were done on a machine with Intel(R) Core(TM) i3-4030U
CPU @ 1.90GHz processor and 4GB RAM.
%
We have compared our performance with the performance of our earlier 
CBMC based implementation.
%
For example, the formula for variation F, the total number of
compartments ($|N|$) equals to 10, returns in 129.78 minutes (7786.8 secs)
with a SAT result.
%
In comparison, CBMC results in OUT OF MEMORY for $|N|$ greater than 5.
%
``!'' indicate that the constraints were unsatisfiable.
%
Hence with the use of this novel encoding, we are able to scale the
system to a much larger compartmentalized system, especially to
eukaryotic cells with a total number of 10 compartments. \textbf{Not only we have improved finding a SAT model but this encoding also did better in comparison to cbmc in the
case of refutation that no model exists. Refer table variant A timing
comparison and VARIANT D with N =2.}
%
% \ankit{
% This additional scalability can provide us leverage to ask for related
% open questions that were not previously possible, for example, what is
% the minimum number of molecules required to satisfy the vesicular
% traffic system in eukaryotes.}
%
Furthermore, we experimented with limits of our tools and found
that $\zthree$ was able to solve the constraints up to $\sim{14-18}$ nodes.

### 2017-07-20 paper/intro_MTedit.tex
add

The analysis of VTSs is a difficult computational problem because of the combinatorial scaling of graph topologies and regulatory rules. Many questions require us to check useful properties over all possible graphs and regulatory rules. Formal verification tools such as model checkers \cite{clarke1996symbolic, biere2003bounded, clarke2008birth, cimatti2000nusmv, holzmann1997model} and SAT solvers \cite{moskewicz2001chaff,een2004extensible} allow us to do this symbolically for graphs of finite size, without enumerating all instances. In a recent work, we employed the model checker CBMC \cite{CKY03, ckl2004} meant to analyze ANSI C programs for studying this problem. VTSs and SNARE regulations were modeled as C code using arrays to represent graphs and Boolean tables. Queries about graph connectivity requirements for different variations of SNARE regulations were modeled as logical assertions to be checked by the model checker. CBMC uses its own built-in encodings to reduce the analysis problem to a Boolean satisfiability problem which is solved using SAT solvers.
%
However, the scalability of our recent approach was limited (up to
VTSs with 6 nodes) due to the encoding used to model VTSs was not
fine-tuned to the structure of the problem.
%
In particular, we needed to encode conditional reachability between
nodes (refer section 3, stability condition), and solved in CBMC using a
combination of non-determinism and partial enumeration that is
inefficient.

>>>>

The analysis of VTSs is a difficult computational problem because of the combinatorial scaling of graph topologies and regulatory rules. Many questions require us to check useful properties over all possible graphs and regulatory rules. Formal verification tools such as model checkers \cite{clarke1996symbolic, biere2003bounded, clarke2008birth, cimatti2000nusmv, holzmann1997model} and SAT solvers \cite{moskewicz2001chaff,een2004extensible} allow us to do this symbolically for graphs of finite size, without enumerating all instances. In a recent work, we employed the model checker CBMC \cite{CKY03, ckl2004} meant to analyze ANSI C programs for studying this problem. VTSs and SNARE regulations were modeled as C code using arrays to represent graphs and Boolean tables. Queries about graph connectivity requirements for different variations of SNARE regulations were modeled as logical assertions to be checked by the model checker. CBMC uses its own built-in encodings to reduce the analysis problem to a Boolean satisfiability problem which is solved using SAT solvers.
%
However, the scalability of our recent approach was limited (up to
VTSs with 6 nodes) due to the encoding used to model VTSs was not completely
fine-tuned to the structure of the problem.
%
In particular, we needed to encode conditional reachability between
nodes (refer section 3, stability condition), and solved in CBMC using a
combination of non-determinism and partial enumeration that is
inefficient.
